Cover Your Basis: Comprehensive Data-Driven Characterization of the Binary Black Hole Population

### 2022-10-11 src/tex/appendicies.tex
fix up appendix and model/prior table

\begin{eqnarray}
\tau_\lambda \sim \Gamma(a, b) \\
p(\bm{c} | \tau_\lambda) \propto \exp \big[ -\frac{1}{2} \tau_\lambda \bm{c}^{\mathrm{T}} \bm{D}_{r}^{\mathrm{T}} \bm{D}_r \bm{c}  \big] e^{-n\bm{c}}
\end{eqnarray}

\noindent Above $\bm{D}_r$ is the order-$r$ difference matrix, of shape $(n-r \times n)$ and $\Gamma(a,b)$ the Gamma distribution with shape parameters, $a$ and $b$. 
This smoothing prior removes the strong dependence on number and location of knots that arises with using splines. The $\tau_\lambda$ controls the ``strength'' of 
smoothing, or the inverse variance of the Gaussian priors on knot differences. When there are a very large number of knots, 
such that your domain is densely populated with basis coefficients, this allows the freedom for the model to find the smoothing 
scale that the data prefers. One downside of this prior specification as noted in \citet{BayesianPSplines} and \citet{Jullion2007RobustSO}, is the dependence on 
one's choice of value for the $b$ hyperparameter in the Gamma distribution for $\tau_\lambda$. The models introduced in \citet{Jullion2007RobustSO} show how 
to alleviate this. W model $\tau_\lambda$ with a mixture model composed of multiple Gamma distributions each with varying values for $b$. 
With this we now have the full prior on $\tau_\lambda$ as:

\begin{align}
    \bm{p} \sim \mathcal{D}(m) \\
    p(\tau_\lambda | \bm{p}) = \sum_i^m p_i \Gamma(1, b_i)
\end{align}

\noindent Above $\mathcal{D}(m)$, denotes a Dirichlet distribution with $m$ degrees of freedom and ones as the concentration parameters, 
and the $b_i$'s are $m$ different b shape parameters for the Gamma distribution.

@@

@@

\begin{table*}[ht!]
    \centering
    \begin{tabular}{|l|l|l|l|}
    \hline
    \textbf{Model} & \textbf{Parameter} & \textbf{Description} & \textbf{Prior} \\ \hline \hline
    \multicolumn{4}{|c|}{\textbf{Primary Mass Model Parameters}} \\ \hline
    \textsc{M-Spline} & $\bm{c}$ & Basis coefficients & $\sim \mathrm{Smooth}(\tau_\lambda, r, n)$ \\ \cline{2-4} 
     & $\tau_\lambda$ & Smoothing Prior Scale & $\sum_i^m p_i \Gamma(1, b_i)$ \\ \cline{2-4}
     & $\bm{p}$ & Hyper-Parameter weights on smoothing scale mixture prior & $\sim \mathcal{D}(m)$ \\ \cline{2-4}
     & $b_m$ & array of different concentration parameters for $\Gamma$ distribution & equal in $[10^{-9},10^{-1}]$ \\ \cline{2-4} 
     & $r$ & order of the difference matrix for smoothing prior & 2 \\ \cline{2-4} 
     & $n$ & number of knots in the basis spline & 50 \\ \cline{2-4} 
     & $m$ & number of mixture components for prior on $\tau_\lambda$ & 24  \\ \hline \hline
    \multicolumn{4}{|c|}{\textbf{Mass Ratio Model Parameters}} \\ \hline
    \textsc{PowerLaw Mass Ratio} & $\beta$ & slope of the mass ratio power law & $\sim \mathcal{N}(0,2)$ \\ \cline{2-4} 
    & $m_\mathrm{min}$ & minimum BH mass & 3\msun  \\ \hline \hline    
    \multicolumn{4}{|c|}{\textbf{Redshift Evolution Model Parameters}} \\ \hline
    \textsc{PowerLaw Redshift} & $\lambda$ & slope of redshift evolution power law $(1+z)^\lambda$ &  $\sim \mathcal{N}(0,2)$ \\ \hline \hline
    \multicolumn{4}{|c|}{\textbf{Spin Distribution Model Parameters}} \\ \hline
    \textsc{M-Spline Magnitude} & $\bm{c}$ &  Basis coefficients & $\sim \mathrm{Smooth}(\tau_\lambda, r, n)$  \\ \cline{2-4} 
    & $\tau_\lambda$ & Smoothing Prior Scale & $\sum_i^m p_i \Gamma(1, 10^{-3})$ \\ \cline{2-4}
    & $r$ & order of the difference matrix for smoothing prior & 2 \\ \cline{2-4} 
    & $n$ & number of knots in the basis spline & 16 \\ \hline
    \textsc{M-Spline Tilt} & $\bm{c}$ &  Basis coefficients & $\sim \mathrm{Smooth}(\tau_\lambda, r, n)$  \\ \cline{2-4} 
    & $\tau_\lambda$ & Smoothing Prior Scale & $\sim \Gamma(1, 10^{-3})$ \\ \cline{2-4}
    & $r$ & order of the difference matrix for smoothing prior & 2 \\ \cline{2-4} 
    & $n$ & number of knots in the basis spline & 16 \\ \hline
    \textsc{M-Spline Effective} & $\bm{c}$ &  Basis coefficients & $\sim \mathrm{Smooth}(\tau_\lambda, r, n)$  \\ \cline{2-4} 
    & $\tau_\lambda$ & Smoothing Prior Scale & $\sim \Gamma(1, 10^{-3})$ \\ \cline{2-4}
    & $r$ & order of the difference matrix for smoothing prior & 2 \\ \cline{2-4} 
    & $n$ & number of knots in the basis spline & 24 \\ \hline
    \end{tabular}
\end{table*}

\bruce{Need to fill this section in and put tables with specific hyperparameter/prior settings made}

%\section{Validation Studies} \label{sec:validation}

%bruce{Need to fill this section}

>>>>

\begin{eqnarray}
\bm{c} \sim \mathcal{N}(0, \sigma) \\
p(\bm{c} | \tau_\lambda) \propto \exp \big[ -\frac{1}{2} \tau_\lambda \bm{c}^{\mathrm{T}} \bm{D}_{r}^{\mathrm{T}} \bm{D}_r \bm{c}  \big] 
\end{eqnarray}

\noindent Above $\bm{D}_r$ is the order-$r$ difference matrix, of shape $(n-r \times n)$, and $\mathcal{N}(0,\sigma)$ a Gaussian distribution with zero mean 
and standard deviation, $\sigma$. This smoothing prior removes the strong dependence on number and location of knots that arises with using splines. 
The $\tau_\lambda$ controls the ``strength'' of the smoothing, or the inverse variance of the Gaussian priors on knot differences. We 
place uniform priors on $\tau_\lambda$ marginalize over this smoothing scale hyperparameter to let the data inform the optimal scale needed.
When there are a very large number of knots, such that your domain is densely populated with basis coefficients, this allows the freedom for the model to find the smoothing 
scale that the data prefers.

%One downside of this prior specification as noted in \citet{BayesianPSplines} and \citet{Jullion2007RobustSO}, is the dependence on 
%one's choice of value for the $b$ hyperparameter in the Gamma distribution for $\tau_\lambda$. The models introduced in \citet{Jullion2007RobustSO} show how 
%to alleviate this. W model $\tau_\lambda$ with a mixture model composed of multiple Gamma distributions each with varying values for $b$. 
%With this we now have the full prior on $\tau_\lambda$ as:

%\begin{align}
%    \bm{p} \sim \mathcal{D}(m) \\
%    p(\tau_\lambda | \bm{p}) = \sum_i^m p_i \Gamma(1, b_i)
%\end{align}

%\noindent Above $\mathcal{D}(m)$, denotes a Dirichlet distribution with $m$ degrees of freedom and ones as the concentration parameters, 
%and the $b_i$'s are $m$ different b shape parameters for the Gamma distribution.

@@

\begin{table*}[hb!]
    \centering
    \begin{tabular}{|l|l|l|l|}
    \hline
    \textbf{Model} & \textbf{Parameter} & \textbf{Description} & \textbf{Prior} \\ \hline \hline
    \multicolumn{4}{|c|}{\textbf{Primary Mass Model Parameters}} \\ \hline
    \textsc{B-Spline} & $\bm{c}$ & Basis coefficients & $\sim \mathrm{Smooth}(\tau_\lambda, r, n)$ \\ \cline{2-4} 
     & $\tau_\lambda$ & Smoothing Prior Scale & U(1,100) \\ \cline{2-4}
     & $r$ & order of the difference matrix for smoothing prior & 2 \\ \cline{2-4} 
     & $n$ & number of knots in the basis spline & 64 \\ \cline{2-4} 
    \multicolumn{4}{|c|}{\textbf{Mass Ratio Model Parameters}} \\ \hline
    \textsc{B-Spline} & $\bm{c}$ & Basis coefficients & $\sim \mathrm{Smooth}(\tau_\lambda, r, n)$ \\ \cline{2-4} 
     & $\tau_\lambda$ & Smoothing Prior Scale & U(1,10) \\ \cline{2-4}
     & $r$ & order of the difference matrix for smoothing prior & 2 \\ \cline{2-4} 
     & $n$ & number of knots in the basis spline & 18 \\ \cline{2-4}  
    \multicolumn{4}{|c|}{\textbf{Redshift Evolution Model Parameters}} \\ \hline
    \textsc{PowerLaw+B-Spline} & $\lambda$ & slope of redshift evolution power law $(1+z)^\lambda$ &  $\sim \mathcal{N}(0,3)$ \\ \cline{2-4}
    & $\bm{c}$ & Basis coefficients & $\sim \mathrm{Smooth}(\tau_\lambda, r, n)$ \\ \cline{2-4} 
     & $\tau_\lambda$ & Smoothing Prior Scale & U(1,10) \\ \cline{2-4}
     & $r$ & order of the difference matrix for smoothing prior & 2 \\ \cline{2-4} 
     & $n$ & number of knots in the basis spline & 16 \\ \cline{2-4} 
    \multicolumn{4}{|c|}{\textbf{Spin Distribution Model Parameters}} \\ \hline
    \textsc{B-Spline Magnitude} & $\bm{c}$ &  Basis coefficients & $\sim \mathrm{Smooth}(\tau_\lambda, r, n)$  \\ \cline{2-4} 
    & $\tau_\lambda$ & Smoothing Prior Scale & U(1,10) \\ \cline{2-4}
    & $r$ & order of the difference matrix for smoothing prior & 2 \\ \cline{2-4} 
    & $n$ & number of knots in the basis spline & 18 \\ \cline{2-4} 
    \textsc{B-Spline Tilt} & $\bm{c}$ &  Basis coefficients & $\sim \mathrm{Smooth}(\tau_\lambda, r, n)$  \\ \cline{2-4} 
    & $\tau_\lambda$ & Smoothing Prior Scale & U(1,10) \\ \cline{2-4}
    & $r$ & order of the difference matrix for smoothing prior & 2 \\ \cline{2-4} 
    & $n$ & number of knots in the basis spline & 16 \\ \cline{2-4} 
    \end{tabular}
    \caption{All hyperparameter prior choices for each of the newly introduced basis spline models from this manuscript. See appendix 
    \ref{sec:basis_splines} and \ref{sec:psplines} for more detailed description of basis spline or smoothing prior parameters.}
    \label{tab:model_priors}
\end{table*}

@@

For each of the distributions with basis spline distributions, we have 2 fixed hyperparameters to specify. The number of degrees of freedom, $n$, and the difference 
penalty order for the smoothing prior, $r$. Additionally, one must choose a prior distribution on the smoothing prior scale hyperparameter, $\tau_\lambda$, which we take to be Uniform. 
For the primary mass distribution we model the log probability with a B-Spline interpolated in $\log(m_1)$ space. We follow a similar scheme for the models in mass ratio and spin, but since 
these parameters have natural hard boundaries we model the log probability with B-Splines interpolated in $q$, $a_i$ or $\cos{\theta_i}$ space. The 
redshift distribution is also lacking of a natural upper boundary, which we mitigate by using a semi parametric approach where B-Splines model modulations to the 
simpler underlying \textsc{PowerlawRedshift} model. We both these B-Spline modulations in log probability interpolated in $\log z$ space. We enumerate each of our specific model hyperparameter 
and prior choices in table \ref{tab:model_priors}.

### 2022-10-11 src/tex/methods.tex
fix up appendix and model/prior table

We parameterize each binaries' masses with the primary (more massive component) mass 
and the mass ratio ($q=\frac{m_2}{m_1}$), bounded in the range from 0 to 1. Furthermore, we model 4 of the 6 total 
spin degrees of freedom of a binary merger with each component spin magnitude $a_1$ and $a_2$, along with the cosine of the tilt angle of each component, 
$\cos{\theta_1}$ and $\cos{\theta_2}$. The tilt angle is defined as the angle between each components' spin vector and the binaries orbital angular momentum vector. 
We assume the polar spin angles are uniformly distributed in the orbital plane. For the primary mass distribution, we model the log probability with a B-Spline interpolated over 
knots linearly spaced in $\log(m_1)$ in the range of support given our observations ($[6.5\msun-100\msun]$), such that $\log(p(m_1 | \bm{c})) \propto B_{k}(\log(m_1) | \bm{c})$, where $B_{k}$ is the 
$k$-th order B-Spline function with basis coefficients, $\bm{c}$. We follow the same procedure for the models in mass ratio and spin distributions with knots spaced linearly across each domain, 
so that we have $\log(p(\theta | \bm{c}_\theta)) \propto B_{k}(\theta | \bm{c}_\theta)$, where $\theta$ can be $q$, $a_1$, $a_2$, $\cos{\theta_1}$ or $\cos{\theta_2}$. 
For the spin magnitude and tilt distributions we construct two versions of the model: first, we model  
each component's distribution is independently and identically distribution (IID), where we have a single B-Spline model and parameters (coefficients) for each binary spin. 
Secondly, we model each component's distribution to be unique, fitting separate sets of coefficients for the B-Spline models of the primary and secondary spin distributions. 
Lastly, we fit a population model on the redshift or luminosity distance distribution of BBHs, assuming a $\Lambda\mathrm{CDM}$ cosmology defined by the parameters 
from the Planck 2015 results \citep{Planck2015}. This defines an analytical mapping between each event's inferred luminosity distance, and it's redshift, which we now use interchangeably. 
We take a semi-parametric approach to model the evolution of the merger rate with redshift, following \citet{Edelman_2022ApJ}, that parameterizes modulations to an underlying model 
with splines (in our case basis Splines). We use the \textsc{PowerlawRedshift} model as the underlying distribution to modulate, which has a single hyperparameter, $\lambda_z$, and 
probability density defined as: $p(z|\lambda_z)\propto \frac{dV_c}{dz}(1+z)^{\lambda_z-1}$ \citep{Fishbach_2018redshift}. For more detailed descriptions of each model and 
specific prior choices used for the hyperparameters fit in Appendix \ref{sec:modelpriors}. Now that we have our comprehensive data-driven population model over 
each of the 7 modeled parameters built, we fit our model using the usual hierarchical Bayesian inference framework (see appendix \ref{sec:hierarchical_inference} for a review) in order to characterize 
the population of BBHs in the most recent catalog of GW observations, GWTC-3 \citep{GWTC3,o3b_astro_dist}.

>>>>

We parameterize each binaries' masses with the primary (more massive component) mass 
and the mass ratio ($q=\frac{m_2}{m_1}$), bounded in the range from 0 to 1. Furthermore, we model 4 of the 6 total 
spin degrees of freedom of a binary merger with each component spin magnitude $a_1$ and $a_2$, along with the cosine of the tilt angle of each component, 
$\cos{\theta_1}$ and $\cos{\theta_2}$. The tilt angle is defined as the angle between each components' spin vector and the binaries orbital angular momentum vector. 
We assume the polar spin angles are uniformly distributed in the orbital plane. For the primary mass distribution, we model the log probability with a B-Spline interpolated over 
knots linearly spaced in $\log(m_1)$ in the range of support given our observations ($[6.5\msun-100\msun]$), such that $\log(p(m_1 | \bm{c})) \propto B_{k}(\log(m_1) | \bm{c})$, where $B_{k}$ is the 
$k$-th order B-Spline function with basis coefficients, $\bm{c}$. We follow the same procedure for the models in mass ratio and spin distributions with knots spaced linearly across each domain, 
so that we have $\log(p(\theta | \bm{c}_\theta)) \propto B_{k}(\theta | \bm{c}_\theta)$, where $\theta$ can be $q$, $a_1$, $a_2$, $\cos{\theta_1}$ or $\cos{\theta_2}$. 
For the spin magnitude and tilt distributions we construct two versions of the model: first, we model  
each component's distribution as independently and identically distribution (IID), where we have a single B-Spline model and parameters (coefficients) for each binary spin. 
Secondly, we model each component's distribution to be unique, fitting separate sets of coefficients for the B-Spline models of the primary and secondary spin distributions. 
Lastly, we fit a population model on the redshift or luminosity distance distribution of BBHs, assuming a $\Lambda\mathrm{CDM}$ cosmology defined by the parameters 
from the Planck 2015 results \citep{Planck2015}. This defines an analytical mapping between each event's inferred luminosity distance, and it's redshift, which we now use interchangeably. 
We take a semi-parametric approach to model the evolution of the merger rate with redshift, following \citet{Edelman_2022ApJ}, that parameterizes modulations to an underlying model 
with splines (in our case basis Splines). We use the \textsc{PowerlawRedshift} model as the underlying distribution to modulate, which has a single hyperparameter, $\lambda_z$, and 
probability density defined as: $p(z|\lambda_z)\propto \frac{dV_c}{dz}(1+z)^{\lambda_z-1}$ \citep{Fishbach_2018redshift}. For more detailed descriptions of each model and 
specific prior choices used for the hyperparameters fit in Appendix \ref{sec:modelpriors}. Now that we have our comprehensive data-driven population model over 
each of the 7 modeled parameters built, we fit our model using the usual hierarchical Bayesian inference framework (see appendix \ref{sec:hierarchical_inference} for a review) in order to characterize 
the population of BBHs in the most recent catalog of GW observations, GWTC-3 \citep{GWTC3,o3b_astro_dist}.
